% -*- Mode: LaTeX -*-
% $Id: LinearAlgebra.tex,v 1.1 1995/01/30 02:02:03 rz Exp $
\chapter{Linear Spaces}
\label{Linear:Chap}

The general approach in Weyl is to use functors to create more complex
domains from simple domains. The simplest such functors are those that
takes a ring or field $R$ and produces a vector space whose elements
lie in $R$. This chapter discusses these domains and the operations
that apply to their elements. The simplest of these domains are free
modules and vector spaces. Both are modules (you can add their
elements and multiply them by elements of $R$). The difference is that
elements of a vector space can also be divided by elements of $R$,
\ie, $R$ is a field. Free modules and vector spaces are discussed in
\sectref{FreeModules:Sec} and \sectref{VectorSpace:Sec} respectively.

\section{Vector Spaces}
\label{VectorSpace:Sec}

Vector spaces are free modules where the coefficient domain is a field.

\begin{functiondef}{get-vector-space}{domain dim}
Create a vector space of dimension {\em dim} where the vectors'
components are all elements of {\em domain}.  {\em Domain} must be a
field.
\end{functiondef}


\section{General Matrices}
\label{GMatrix:Sec}

\begin{functiondef}{get-matrix-space}{ring}
Returns a general matrix domain, where the entries in the matrices are
elements of {\em ring\/}.
\end{functiondef}

\begin{functiondef}{make-element}{domain value \rest values}
Make an element of the module {\em domain}, where the first component
is {\em value}, etc.  
\end{functiondef}

The following two functions return square matrices whose off diagonal
elements are all zero. Its diagonal elements are $0$ for
\keylisp{zero-matrix} and $1$ for \keylisp{one-matrix}.

\begin{flushleft}
\functdef{zero-matrix}{matrix-domain \optional rank} \\
\functdef{one-matrix}{matrix-domain \optional rank}
\end{flushleft}

\begin{functiondef}{matrix-dimensions}{matrix}
Returns the dimensions of this particular matrix.
\end{functiondef}

\begin{functiondef}{ref}{vector i j}
Returns the $(i,j)$ element (zero based) of {\em vector\/}.  If $i$ or
$j$ is {\sf :*}, then a row or column vector is returned.
\end{functiondef}

\begin{functiondef}{with-matrix-dimensions}{(dim1 dim2 \optional array)
matrix \body body}
This control abstraction binds {\em dim1} and {\em dim2} to the
dimensions of {\em matrix}.  If {\em array} is provided, it is bound
to an array that whose elements are the elements of the {\em
matrix\/}.  This control structure can lead to substantially more
efficient manipulation of dense matrices.  

Ultimately, {\em array} will turn into an efficient accessor for the
elements of {\em matrix} so that this technique will work with other
than dense matrices.
\end{functiondef}

\begin{functiondef}{transpose}{matrix}
Transpose {\em matrix\/}.
\end{functiondef}

\begin{functiondef}{jacobian}{function-list var-list}
Returns the Jacobian of {\em function-list} with respect to {\em
vector-list}.  These should really be vectors, but list will work
also.   
\end{functiondef}

\begin{functiondef}{direct-sum}{\rest matrices}
Assumes the matrices all have the same
number of rows, $n$.  It returns a matrix of $n$ rows, where
each row is the concatenation of the rows of each of
its arguments.
\end{functiondef}

\begin{functiondef}{determinant}{matrix}
Returns the determinant of the {\em matrix}.
\end{functiondef}
  
\begin{functiondef}{sparse-determinant}{matrix}
Returns the determinant of the {\em matrix}.
\end{functiondef}
  
\begin{functiondef}{subdeterminant}{matrix}
Returns the determinant of a nonsingular submatrix of the {\em matrix}
of maximum rank.
\end{functiondef}
  
\begin{functiondef}{hermite}{matrix}
Returns the \keyi{Hermite normal} form of the {\em matrix}.
At the moment, the method is implemented only for matrices
over the ring of integers.
\end{functiondef}
  
\begin{functiondef}{smith}{matrix}
Returns the vector of diagonal elements of the \keyi{Smith normal
form} of the {\em matrix}.  At the moment, the method is implemented
only for matrices over the ring of integers.
\end{functiondef}
  
\section{Matrix Groups}





